Finally, it is time to write this hard part of the thesis. The acknowledgments! Although it is an optional element of the document, I think that it should be mandatory. It is nearly impossible to imagine developing a PhD thesis alone. Many people work together and backstage, but all are fundamental to reach the final objective. Now it is time to thank them all.

\vspace*{0.5cm}

Firstly, I would like to thank my advisor team. Indeed, I do not have just an advisor and a co-advisor, but a team of three advisors. Each one started in a distinct phase of this thesis. So, I thank you, Prof. Dr. André Du Bois, for accepting me as a PhD student. Also, for your support and confidence in me. Prof. Dr. Maurício Pilla, for sharing profound knowledge and guidance during my initial studies on the thesis subject. Finally, Dr. Matthias Diener, I have no words to thank you for all your support. Our contact started with a simple question about your work on Facebook and finished with you being officially my co-advisor. My sincerely \textit{Herzlichen Dank}.

\vspace*{0.5cm}

I also would like to thank another team that was always working backstage: my family! Especially my wife Daiane, my son Nicolas and my mom Maria. Thank you for the support and encouragement during these four long years, including giving me the strengths to continue when sometimes I thought that I would not be able to make it! But I did it! This achievement is also yours.

\vspace*{0.5cm}

Lastly, I would like to thank my employer (CPD and UFSM) to grant me the privilege of getting a license of my professional activities. Hence, it was possible to dedicate my full time to my PhD studies.

\vfill
\textit{This study was financed in part by the Coordena\c{c}\~ao de Aperfei\c{c}oamento de Pessoal de N\'ivel Superior - Brasil (CAPES) - Finance Code 001 and PROCAD/LEAPaD}
